---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(toxcalc)
```

# toxcalc

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![license](https://img.shields.io/badge/license-GPL%20%28%3E=%202%29-lightgrey.svg)](https://choosealicense.com/)
<!-- badges: end -->

## Overview

`toxcalc` is an open-source implementation of the statistical methods specified
in the US EPA Whole Effluent Toxicity (WET) method manuals. It recreates the
capability of ToxCalc v5.0 (Tidepool Scientific), commercial software that has
been withdrawn — the vendor product page now returns a 404.

The package provides both branches of the EPA analysis:

- the **hypothesis-testing** flowchart, giving the no-observed-effect
  concentration (NOEC), the lowest-observed-effect concentration (LOEC), the
  minimum significant difference (MSD) and the percent MSD (PMSD);
- the **point-estimation** flowchart, giving the median lethal concentration
  (LC50) by the graphical, Spearman-Karber, trimmed Spearman-Karber or probit
  method, and the inhibition concentration (ICp) by linear interpolation with
  bootstrap confidence limits.

## Why this exists

The EPA manuals specify not only the statistical tests but the decision rules
that select between them. Those rules are the part that is hardest to reproduce
by hand and easiest to leave undocumented.

Every `toxcalc` analysis therefore carries an **audit trail**: a table recording
each branch point of the flowchart — the question asked, the statistic, the
answer, the consequence, and the section of the EPA manual that justifies it.

```{r}
#| label: audit-trail
summary(toxcalc(fathead_c1, response = "weight",
                pmsd_bounds = "fathead_growth"))
```

Those are the growth data from Appendix C of the chronic manual, and the
flowchart arrives at Dunnett's procedure, the NOEC of 128 and the LOEC of 256
that the manual itself reports.

## Running both branches

A laboratory report usually carries a no-effect concentration and a point
estimate together. `branch = "both"` returns them from one call, and
`as.data.frame()` gives one tidy row per endpoint.

```{r}
#| label: both-branches
fit <- toxcalc(ceriodaphnia_m1, response = "young",
               branch = "both", seed = 42)
as.data.frame(fit)[, c("endpoint", "value", "lower", "upper", "method")]
```

The point estimate uses all the data, including any concentration excluded from
the no-effect concentration, which is what section 9.5.2 requires.

## Point estimation

`lc50()` walks the Figure 6 chart of the acute manual. Each of the four columns
of its Table 20 was constructed to exercise one of the four methods, and each
routes to the right one:

```{r}
#| label: lc50-routing
for (column in c("graphical", "spearman_karber", "trimmed", "probit")) {
  chosen <- lc50(acute_table20, response = column,
                 n_exposed = "exposed", type = "quantal")
  cat(format(column, width = 16), "->", chosen$selected, "\n")
}
```

## Fidelity

The target is faithfulness to the EPA method manuals, validated against the
worked examples printed in them. It is **not** a bit-for-bit clone of ToxCalc;
no ToxCalc output was available for comparison.

Almost every published value reproduces exactly. The exceptions are documented
rather than smoothed over, and each is reconciled by a test:

- **seven printed values do not follow from the manuals' own data**, all through
  hand-calculation rounding or arithmetic slips, and one dataset is not the
  dataset its own text says it is;
- **the trimmed Spearman-Karber interval** is the one quantity that differs for
  a reason other than an error in the manual: the manual delegates it to a
  program whose formula it never states, and Hamilton's corrected variance is
  not publicly retrievable;
- **the ICp interval** cannot be matched because the program that produced it
  names the random seed it used.

The vignette `vignettes/recreating-toxcalc.qmd` records every point at which the
source material was ambiguous, internally inconsistent, or at odds with modern
practice, and what was decided in each case, with the evidence.

`williams()` implements the one ToxCalc feature that is not an EPA method.
Neither manual gives a worked example or a table of critical values for it, so
the values are simulated and the implementation is validated against
mathematical identities instead. It is labelled as an extension throughout, and
neither flowchart will ever select it.

## Status

**Experimental, under active development.** Both branches are complete and
`R CMD check` is clean. Still to come: the marine manual's endpoints, designs
with more than one control, and the remaining alternative tests.

See `notes/TOXCALC-human.md` for the plan and `notes/TOXCALC-claude.md` for the
specification.

## Installation

```r
if (!requireNamespace("remotes")) {
  install.packages("remotes")
}
remotes::install_github("beckyfisher/toxcalc")
```

## References

- US EPA (2002) *Methods for Measuring the Acute Toxicity of Effluents and
  Receiving Waters to Freshwater and Marine Organisms*, 5th ed.
  EPA-821-R-02-012.
- US EPA (2002) *Short-term Methods for Estimating the Chronic Toxicity of
  Effluents and Receiving Waters to Freshwater Organisms*, 4th ed.
  EPA-821-R-02-013.
- US EPA (2002) *Short-term Methods for Estimating the Chronic Toxicity of
  Effluents and Receiving Waters to Marine and Estuarine Organisms*, 3rd ed.
  EPA-821-R-02-014.
- Hamilton MA, Russo RC, Thurston RV (1977) Trimmed Spearman-Karber method for
  estimating median lethal concentrations in toxicity bioassays.
  *Environmental Science & Technology* 11:714-719.
- Norberg-King TJ (1993) *A linear interpolation method for sublethal toxicity:
  the inhibition concentration (ICp) approach*, version 2.0. EPA/600/M-91/037.
- Williams DA (1971) A test for differences between treatment means when several
  dose levels are compared with a zero dose control. *Biometrics* 27:103-117.
