---
title: "Recreating ToxCalc: methods and decisions"
vignette: >
  %\VignetteIndexEntry{Recreating ToxCalc: methods and decisions}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

```{r}
#| label: setup
#| include: false
library(toxcalc)
```

## Purpose

This vignette records how `toxcalc` was built and, more importantly, **every
point at which the source material was ambiguous, internally inconsistent, or
at odds with modern practice, and what was decided in each case**.

The package recreates ToxCalc v5.0 (Tidepool Scientific), commercial software
implementing the US EPA Whole Effluent Toxicity methods, which has been
withdrawn. No ToxCalc output was available for comparison, so the specification
is taken from the EPA method manuals themselves and validated against the
worked examples printed in them.

Every decision below is stated with the evidence that settled it. Where the
package departs from a figure printed in a manual, the departure is
demonstrated, not asserted: each is reconciled by a test in the package test
suite, so the discrepancy cannot later be mistaken for an implementation error.

## Sources

- EPA-821-R-02-012, *Methods for Measuring the Acute Toxicity of Effluents and
  Receiving Waters to Freshwater and Marine Organisms*, 5th edition, October
  2002.
- EPA-821-R-02-013, *Short-term Methods for Estimating the Chronic Toxicity of
  Effluents and Receiving Waters to Freshwater Organisms*, 4th edition, October
  2002. Section references below are to this manual unless stated otherwise.

Both are US Government works and are publicly available.

## The decisions

### 1. Statistical tables are computed, not transcribed

**Decision.** No table of critical values reproduced in the EPA manuals from a
third-party source is copied into this package. Critical values are computed.

**Reasoning.** The manuals are US Government works, but several tables *within*
them are reproductions credited to other authors: Dunnett's Table C.5 to Miller
(1981), the Shapiro-Wilk coefficients of Table B.4 to Conover (1980), and
Williams' critical values to *Biometrics*. Permission to reproduce a table in a
government manual does not transfer to a third party redistributing it.

Two further considerations point the same way. Computation covers designs the
tables never did, in particular unbalanced replication. And the reproduction in
the source document is not always reliable: the Dunnett table as printed
contains values corrupted in transcription, including 1.86 at 120 degrees of
freedom with one treatment, where the sequence requires a value near 1.66.

Only tables EPA generated itself are shipped: the Kolmogorov critical values of
Table B.11, and the variability criteria of Table 6, which EPA derived from its
own interlaboratory study.

### 2. Dunnett's critical value is integrated, not simulated

**Decision.** The critical value is obtained by numerical integration, using
`stats::integrate` and `stats::uniroot`. The obvious alternative,
`mvtnorm::qmvt()`, is not used, and `mvtnorm` is not a dependency.

**Reasoning.** `qmvt()` uses randomised quasi-Monte Carlo integration and
returns a different answer on every call. For the Appendix C design it returned
values between 2.3552 and 2.3572 across repeated calls. An analysis submitted
to a regulator must give the same answer every time it is run.

Because all the comparisons share the same control, the correlation matrix has
a one-factor form: the correlation between comparisons `i` and `j` is
`lambda_i * lambda_j`, where `lambda_i = sqrt(n_i / (n_control + n_i))`.
Conditioning on that single common factor, and on the pooled standard
deviation, reduces the multivariate t probability to a two-dimensional
integral, which ordinary quadrature evaluates to machine precision.

The result agrees with the manual's table and is stable:

```{r}
#| label: dunnett
fit <- msd(fathead_c1, response = "weight")
fit$critical           # EPA Table C.5 prints 2.36
replicate(3, msd(fathead_c1, response = "weight")$critical)
```

### 3. Shapiro-Wilk follows Royston, not the manual's tables

**Decision.** Normality is tested with `stats::shapiro.test()`, which uses
Royston's AS R94 algorithm, and the p-value is compared with the assumption
significance level. The manual's procedure — looking up Conover's coefficients
in Table B.4 and comparing the statistic with a quantile in Table B.6 — is not
implemented.

**Reasoning.** The two agree. On the manual's own Appendix B worked example
Royston returns 0.9601 against a printed 0.959, and the difference is entirely
accounted for by the manual centring its observations with concentration means
rounded to three decimal places; applying the same rounding gives 0.9594.

```{r}
#| label: shapiro
epa_normality(fathead_b1, response = "weight")
```

Royston's algorithm is preferred because it is not limited to the tabulated
sample sizes, it returns a p-value rather than a decision at one fixed cutoff,
and it avoids reproducing a third-party table. The manual's tabulated route is
a hand-calculation aid from before this was available in software.

**Residual risk.** In a borderline case the two could send an analysis down
different branches of the flowchart. The package reports both the statistic and
the p-value so that such a case is visible.

### 4. Three printed statistics are not reproducible from the manual's own data

The manuals were prepared for hand calculation, and several worked examples
carry intermediate values forward after rounding them. Recomputing at full
precision therefore does not always return the printed figure. In every case
found so far the conclusion is unchanged, but the numbers differ, and anyone
checking this package against the manual needs to know which is which.

#### Bartlett's test: 7.691 against 6.836

The Appendix B example prints `B = 7.691`. From the raw data the statistic is
6.836. The manual computed it from its own printed table of variances rounded
to four decimal places, and then divided by a correction factor also rounded,
to 1.133.

```{r}
#| label: bartlett
epa_variance(fathead_b1, response = "weight")$statistic

# reproducing the printed value from the manual's rounded intermediates
printed <- c(0.0018, 0.0020, 0.0001, 0.0059, 0.0037)
nu <- rep(3, 5)
(sum(nu) * log(sum(nu * printed) / sum(nu)) - sum(nu * log(printed))) / 1.133
```

One of those rounded variances is itself wrong: the variance at 32 µg/L is
0.002055, which the manual prints as 0.0020 rather than 0.0021. Bartlett's
statistic depends on the logarithm of each variance, so a small relative error
in the smallest variance moves it appreciably.

**Decision.** Compute from the unrounded data. Both values are far below the
13.277 critical value.

#### Kolmogorov D: 0.4684 against 0.4572

The Appendix B example prints `D* = 0.4684`. At full precision it is 0.4572.
The manual standardises each residual and then looks the probability up in a
printed normal table, which requires rounding each `z` to two decimal places
first. Doing the same reproduces the printed value:

```{r}
#| label: kolmogorov
epa_normality(ceriodaphnia_b7, response = "young")$statistic

residuals <- ceriodaphnia_b7$young -
  ave(ceriodaphnia_b7$young, ceriodaphnia_b7$conc,
      FUN = function(z) round(mean(z), 1))
n <- length(residuals)
p <- pnorm(round(sort(residuals) / sd(residuals), 2))
d <- max(max(seq_len(n) / n - p), max(p - (seq_len(n) - 1) / n))
d * (sqrt(n) - 0.01 + 0.85 / sqrt(n))
```

**Decision.** Compute at full precision. Both values are far below the 1.035
critical value.

#### Appendix C section 1.11: 0.087 against 0.162

The same section that computes `MSD = 0.162` then describes the minimum
detectable difference as 0.087 mg, "a decrease in growth of 24 % from the
control". The control mean is 0.677, and 24 % of it is 0.162, so 0.162 is
correct and 0.087 is a typographical error.

```{r}
#| label: msd
pmsd(fathead_c1, response = "weight", bounds = "fathead_growth")
```

### 5. Appendix C does not use the data it says it does

Appendix C states that its data are "the same data used in Appendices B and
D". Table C.1 differs from Table B.1 at fifteen of twenty values. The two are
shipped as separate datasets, `fathead_b1` and `fathead_c1`, and each is
checked against the summary statistics printed beside it.

```{r}
#| label: datasets
rbind(
  `Table B.1` = tapply(fathead_b1$weight, fathead_b1$conc, mean),
  `Table C.1` = tapply(fathead_c1$weight, fathead_c1$conc, mean)
)
```

Using the wrong one is not a small matter: Table B.1 gives a control mean of
0.714 and a within mean square root of 0.052, against 0.677 and 0.097 for Table
C.1, and the two lead to different no-observed-effect concentrations.

### 6. Smoothing is full pool-adjacent-violators, not pairwise averaging

**Decision.** `smooth_monotone()` implements the pool-adjacent-violators
algorithm, with block sizes tracked so that a run of several values pools to
their common mean in one step.

**Reasoning.** The manuals describe smoothing as averaging a violating *pair*.
Taken literally that is wrong, because their own worked examples pool longer
runs: the Appendix K mortality proportions collapse from five values to a
single 0.02, which repeated pairwise averaging does not produce. The manuals
acknowledge this obliquely, noting that "unusual patterns in the deviations
from monotonicity may require an additional step of smoothing".

```{r}
#| label: smoothing
smooth_monotone(c(0.05, 0.00, 0.05, 0.00, 0.00), direction = "increasing")
```

Weights are exposed because the two places the manuals use smoothing need
different ones. The linear interpolation method smooths concentration means
with equal weight, whereas the isotonic means of Williams' test are weighted by
replication.

### 7. The arc sine transformation includes its endpoint adjustments

**Decision.** `arcsine_sqrt()` substitutes `1 / (4 * n)` for an observed proportion
of 0 and `1 - 1 / (4 * n)` for a proportion of 1, where `n` is the number of
organisms in the replicate, and requires `n` whenever either occurs.

**Reasoning.** These adjustments are part of the method, not a refinement. A
replicate at 0 or 1 would otherwise contribute no variance and the tests
downstream would understate the true variability. They reproduce the manual's
printed values exactly:

```{r}
#| label: arcsine
c(
  `RP = 0.60` = arcsine_sqrt(0.60),   # manual prints 0.8861
  `RP = 0`    = arcsine_sqrt(0, n = 20),  # manual prints 0.1120
  `RP = 1`    = arcsine_sqrt(1, n = 20)   # manual prints 1.4588
)
```

### 8. Williams' test is an extension, not an EPA method

**Decision.** Williams' test will be implemented, because ToxCalc offered it,
but it is documented as an extension and is not on the flowchart the package
walks by default.

**Reasoning.** The manuals do not include Williams' test in the
hypothesis-testing flowchart of Figure 2. Section 9.4.1.2 mentions it only as
an alternative requiring additional assumptions, principally that the
concentration-response relationship is monotone. Presenting it as an EPA method
would misrepresent the manuals.

### 9. Two significance levels, not one

**Decision.** The package carries `alpha_assumption`, defaulting to 0.01, and
`alpha`, defaulting to 0.05.

**Reasoning.** Section 9.4.6.1 sets the assumption tests at 0.01 and the
hypothesis tests at 0.05, one-sided. A single significance level argument would
be wrong for one or the other.

### 10. Dependencies are kept to what is genuinely needed

**Decision.** At runtime the package depends on `chk` and base `stats` only.
`MASS`, `multcomp`, `mvtnorm`, `nortest` and `boot` appear in `Suggests`, where
they are used in the test suite as independent checks on methods implemented
here from first principles.

**Reasoning.** Each was considered and rejected as a runtime dependency.
`mvtnorm` was displaced by the integration in decision 2. `multcomp` would have
been used for one quantile while pulling in four further packages. `nortest`
provides statistics the manuals do not specify. `MASS::dose.p` supplies a
median lethal concentration but none of the Abbott adjustment, heterogeneity
test or Fieller limits the manuals also require. `boot` cannot express the EPA
bootstrap, which resamples within concentration, re-smooths inside every
iteration, and takes an order-statistic interval.

`PMCMRplus`, the usual source for Williams' and Steel's tests, is not used: it
requires the external MPFR and GMP libraries, which makes it impractical to
install on Windows.

## Still open

These are recorded rather than resolved, and will be settled against the
published worked examples when the phases that need them are implemented.

- **The trimmed Spearman-Karber confidence interval.** Neither the multiplier
  (1.96 or 2) nor the variance denominator (`n_i` or `n_i - 1`) is stated. The
  two published examples have enough precision to determine both.
- **The probit interval.** Whether proportions are Abbott-adjusted before
  fitting or a natural-response parameter is fitted, whether limits are
  Fieller's or the delta method, and whether a heterogeneity factor is applied,
  are all unstated and interact.
- **The expanded interval for fewer than seven replicates**, which is defined
  only in the ICPIN v2.0 program documentation and is not publicly retrievable.
  The current intention is to omit it and report the ordinary interval with a
  note.
- **Non-monotone patterns of significance.** Section 9.6.5.1 advises caution
  where a lower concentration is not significant between two that are, but
  prescribes no rule.

## Checking this package against the manuals

Every claim above is a test. To run them:

```r
devtools::test()
```

The datasets are rebuilt by `data-raw/DATASET.R`, which re-checks each
transcription against the summary statistics printed beside it in the manual
and fails if any value has drifted.
